How Slow do my Transient Typechecks Go?

### 2019-08-17 paper.tex
Horrible section 5

Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained (over the mean time after warmup) for any of experiments was $\pm3.53\%$ (for the PyStone benchmark).

All experiments were executed on a Windows 10 machine
running Fedora 30 within Windows Subsystems for Linux (version 1). The machine has 
an Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
Benchmarks were executed one by one to avoid interference between them.
The analysis of the results and plots where generated using Python 3.7.3 and PGFPLOTS 1.16.
To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, are available online.\footnote{
\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}

@@

\begin{figure*}
	\input{graphs/full-scatter}
	\caption{Graphs of (at most) 102 configurations in the typing latices for each benchmark. Time is measured as the mean of the 351\textsuperscript{st} to the 1,000\textsuperscript{th} benchmark iteration under a single invocation of Moth (lower is better).}
	\label{f:full}
\end{figure*}
The results for the performance of our sample of typing lattice configurations are presented in figure~\ref{f:full}. Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution of each individual configuration. The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations respectively. The execution time in milliseconds is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.

Most of the graphs are essentially horizontal lines, indicating that the overhead of including type annotations is negligible. The plots for CD and Richards show a roughly linear increase, however, i.e.\ for these two benchmarks, adding type annotations increases performance linearly. On the other hand, the plots for Go, Permute, DeltaBlue, and Storage show \emph{decreases}: i.e. adding more type annotations \emph{increases} performance of these benchmarks.

Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly horizontal lines. Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away. List appears to have three performance modes: most of the configurations up to about 60\% annotations perform at roughly the same speed as untyped code; most configurations above 70\% perform 1.8 times more slowly than untyped code; and then across all annotation percentages there are configurations that perform roughly 1.5 times slower than the untyped baseline.

\section{Individual Annotations}
\begin{figure*}
	\input{graphs/pattern}
	\caption{Graphs of typing latices and performance with only an individual type annotation.}
	\label{f:pattern}		
\end{figure*}
Figures

We hypothesise that the tri/bi-modal graphs shown in figure~\ref{f:full}, are caused by individual type annotations: i.e. certain annotations have a significant effect on performance. In order to investigate this, we ran further experiments by testing configurations with each \emph{individual} type annotation enable, and all others removed, we then used this information to approximate and see if this appears to show a relation with the results across the typing lattice.

Firgure X shows the patterns we were able to identify, in particular...

\label{s-individual}

We hypothesised that the overhead of an individual type annotation may
be able to predict the performance of program configurations with that
annotation. To test this hypothesis, we constructed a series of
configurations, one for each type annotation, with only that
one annotation included.

   trying-to-be-cleverer results

   - single annotation overheads
  - using them go back to scatterplots (i.e. what you do in the latest graphs)

  - show that single annotation prediction doesn't really work\ldots

>>>>

Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained (over the mean time after warmup) for any of experiments was $\pm3.36\%$ (for the PyStone benchmark).

All experiments were executed on a Windows 10 machine
running Fedora 30 within Windows Subsystems for Linux (version 1). The machine has an Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. This computer appears to produce comparable results to the one used in our previous work, although the absolute performance is higher, the relative overhead of transient type-checks is similar.

We used the same version of Moth as previously, together with ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43. Benchmarks were executed one by one to avoid interference between them. The analysis of the results and plots where generated using Python 3.7.3 and PGFPLOTS 1.16. To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, are available online.\footnote{\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}

@@

\begin{figure*}
	\input{graphs/full-scatter}
	\caption{Graphs of (at most) 102 configurations in the typing latices for each benchmark. Time is measured as the mean of the 351\textsuperscript{st} to the 1,000\textsuperscript{th} benchmark iteration under a single invocation of Moth (lower is better).}
	\label{f:full}
\end{figure*}
The results for the performance of our sample of typing lattice configurations are presented in figure~\ref{f:full}. Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution if each individual configuration. The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations (respective). The execution time (in milliseconds) is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.

Most of the graphs are mostly horizontal lines: indicating that the performance effect of including type annotations is negligible. However the plots for CD and Richards show a roughly linear increase: i.e. adding more type annotations to the program increases performance linearly. The plots for Go, Permute, DeltaBlue, and Storage even show \emph{decreases}: i.e. adding more type annotations can actually \emph{increase} performance. Of particular note is that some of the graphs appear to show bimodal performance profiles (Permute, Storage, Snake, Towers, and Richards), or in the case of List, trimodal.

\section{Individual Annotations}
\label{s-individual}
\begin{figure*}
	\input{graphs/pattern}
	\caption{Pairs of colour coded scatter and column graphs. The scatter graphs represent the performance of a sample of typing lattice. The column graphs show the performance of every configuration with only one type annotation. The scatter plots and columns graphs are colour coded based on whether a particular type annotation or two are present in the source code.}
	\label{f:pattern}		
\end{figure*}
We hypothesise that the tri/bi-modal graphs shown in figure~\ref{f:full}, are caused by individual type annotations: i.e. certain annotations have a significant effect on performance. In order to identify such type annotations, for each benchmark and type annotation, we manually inspected the scatter plots for the benchmark's typing lattice, coloured based on whether the chosen type annotation was present or not. This resulted in 1,775 graphs, however most of these showed no discernible pattern.

As an alternative, and less labour intensive approach to identifying potentially influential type annotations, for each type annotation we ran each benchmark with only that type annotation present. We used the same computer and number of iterations as for the typing lattice experiments, albeit these were done on different days.

Figure \ref{f:pattern} shows the results of these experiments, as well as the patterns we manually identified.

The column graphs show the execution time of configurations with with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the 1\textsubscript{st} annotation, and the last column represents the last one (in the order they appear in the source code). The y-axes for the column graphs are the same as the associated scatter graph. However, though the right y-axes for the column graphs appear to be identical the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments where run.

For each benchmark we highlighted one or two type annotations that appear to show a pattern for the typing lattice performance, or in the case of SpectralNorm and Json had higher than usual columns. The identified types are those represented by the red and blue columns. The scatter plots are colour coded accordingly: red/blue circle represents those with the given type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.

As can bee seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs. For the Snake, Towers and CD benchmark, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs. For permute there is actually a signficant performance \emph{incrase}, indicated by the lower half of the lattice being highlighted in red. DeltaBlue and Go also show a slight incrase in performance, but here this is caused by two type annotations (in red and blue), however this does not appear to be comulative (as indicated by the purple dots being about the same height as the red and blue ones).

List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present. Richards is particularly od as it shows a performance decrease that appears to occur when the red type annotations is present, but not the blue one.

Finally, SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a pattern in the typing lattice.

Note however that in the case of DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively), however we found no apparent relation between these and the typing lattice.

The remaining benchmarks, which we have not shown in \ref{f:pattern}, have relatively flat column graphs and we could not identify any patterns in the typing lattice. The benchmark with the greatest difference in single type configurations relative to the untyped configuration is Storage, with a configuration only $7.9\%$ faster than the untyped baseline. However, though Storage (as shown in \ref{f:full}) appears to be bimodal like Towers, we were unable to find any relationships between its typing lattice and any individual type annotations.
%For example for Snake, the dots covered in red (corresponding to configurations with type annotation number 61) are noticeably higher than the grey ones (those without type annotation 61), thus it appears that the 61\textsuperscript{st} type annotation causes a noticeable slowdown, which is also exemplified by the higher than average red column. The graphs for CD (type \#107), Towers (type \#7) show similar patterns. The graph for Permute is similar, but it shows that type annotation 8 actually makes the benchmarks \emph{faster}.
%In some cases, more than two type annotations have an effect, for DeltaBlue and Go it appears that types 154/155, and 31/47 (respectively) all cause an increase in performance, but this is not cumulative.
%However with List, the performance identified by type annotations 17 and 18 is actually comulative

%For most of the graphs, the identified type annotations are the same, however this is not always the case. For SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), there appears to be no relationship with the typing lattice.
%For DeltaBlue and Richards, type annotations 62 and 24 on the column graphs appear to be significant, however we did not see any kind of pattern their.
